\documentclass[a4paper, 10pt]{article}
\topmargin-4.0cm

\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage[a4paper, margin=2cm]{geometry}
\usepackage{fancyhdr}
\usepackage{color}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{hyperref}
\usepackage{lastpage}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}
\usepackage{datetime}

\usepackage{newpxtext}
\usepackage{newpxmath}

\begin{document}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

\newdate{date}{17}{03}{2020}
\newdate{friday}{20}{03}{2020}
\newdate{next}{24}{03}{2020}

\begin{center}
{\LARGE \bf Meeting Notes}\\
\displaydate{date}
\end{center}

\section{Talk of the day}

\begin{itemize}
\item A\"ina has finished proving a version of the awkward example without the preamble and with stack clearing upon return.

\item Alix, Arma\"el and Thomas have finished the refactoring.

\item Alix has rebased the uninitialized capabilities branch over the refactoring and started fixing proofs.

\item Thomas has given a presentation of what the driver code that would be verified could look like.
\end{itemize}

\section{Things to do for next meeting}

\begin{itemize}
\item A\"ina will give a tour of the Coq code on \displaydate{friday} to showcase how to go through the examples. She will also give a higher level presentation of the awkward example, and how to reason about the world/revocation in the Iris setting next meeting.

\item Arma\"el and Thomas will write down something about the operational semantics for I/O, and the expected changes to the value relation.

\item Alix will write down something about uninitialized capabilities, how they are modeled, what is impacted by their introduction, and try to continue working on the uninitialized capabilities branch.
\end{itemize}


\end{document}
